Investigating soundscapes perception through acoustic scenes simulation.

### 2017-09-10 lafaySimulatedScenePerception.tex
add new corr value, add figure XP2, samll correction

\newcommand{\gl}[1]{\textcolor{magenta}{GL : #1}}
\newcommand{\nm}[1]{\textcolor{magenta}{#1}}
\newcommand{\ml}[1]{\textcolor{red}{#1}}
\newcommand{\jfp}[1]{\textcolor{blue}{JFP : #1}}
\newcommand{\ie}{\emph{i.\,e.}}
\newcommand{\Ie}{\emph{I.\,e.}}
\newcommand{\cf}{cf.}
\newcommand{\Cf}{Cf.}
\newcommand{\vs}{\emph{vs.}}
\newcommand{\Vs}{\emph{Vs.}}
\newcommand{\al}{\emph{et~al.}}
\newcommand{\eg}{\emph{e.\,g.}}
\newcommand{\Eg}{\emph{E.\,g.}}
\newcommand{\etc}{\emph{etc.}}
\newcommand{\myfloatalign}{\centering}

@@

  \begin{abstract}
    This paper introduces a new experimental protocol to study mental representations of urban soundscapes through a simulation process. Subjects are asked to create a full soundscape by means of a dedicated software tool, coupled with a structured sound data set. This paradigm is used to characterize urban sound environment representations by analyzing the sound classes that were used to simulate the auditory scenes. A rating experiment of the soundscape pleasantness using a 7-points Likert scale is conducted to further refine the analysis of the simulated urban acoustic scenes. Results show that 1) a semantic characterization in terms of presence~/ absence of sound sources is an effective way to characterize urban soundscapes pleasantness, and 2) physical descriptors computed for specific sound sources better characterize the appraisal than global descriptors.
  \end{abstract}

  \begin{keyword}
  %% keywords here, in the form: keyword \sep keyword
 cognitive psychology \sep soundscape perception \ soundscape simulator
  %% PACS codes here, in the form: \PACS code \sep code

@@

\begin{figure}
        \myfloatalign
        \includegraphics[trim={0 0 0 2.5cm},clip,width=.8\linewidth]{gfx/ch_4/3.png}
       \caption{Hierarchical organisation of the isolated sounds used in the simulation.}\label{fig:orgDb}
\end{figure}

@@

\begin{figure}
		\centering
        \subfloat[]
        {\includegraphics[width=.8\linewidth]{gfx/ch_4/SSF}\label{fig:ssf}} \par
        \subfloat[]
        {\includegraphics[width=\linewidth]{gfx/ch_4/simscene}\label{fig:simscene}}
       \caption{Graphical interfaces for the selection of sound classes (a) and their sequencing (b) of the \emph{simScene} tool.}
\end{figure}

@@

\begin{figure}
        \includegraphics[width=\linewidth]{gfx/ch_5/5.png}
        \caption{Experimental protocol of the simulation experiment (1.a) and the pleasantness evaluation experiment (1.b)}\label{fig:xp1_2}
\end{figure}

@@

44 students (14 females; averaging 21.6 years of age, s.d. of 2.0 years) from \emph{Ecole Centrale de Nantes} (a french engineering school) took part to the experiment. They are all living in downtown Nantes (a french city) since 2 years or more.

@@

The assessment is done with a 7-point Likert scale going from -3 (non ideal / unpleasant) to +3 (ideal / pleasant). Before evaluating a scene, participants must listen to the first 20 seconds of the stimuli. After the evaluation, they are free to switch to the next scene.

@@

10 students (2 females; averaging 23.1 years of age, s.d. of 1.8 years) from \emph{Ecole Centrale de Nantes} (a french engineering school) took part to the experiment. They are living in downtown Nantes (a french city) since 2 years or more. None of them took part to the previous simulation experiment (Exp. 1.a).

@@

\begin{table}[t]
\centering
\begin{tabular}{c c}
Terms                         & Acronyms              \\
\hline
sound Level                        & $L$                    \\
sound Level (events)            & $L(E)$                 \\
sound Level (textures)              & $L(T)$                 \\
Average pleasantness (per scene)     & $\mathcal{A}_{scene}$  \\
Average pleasantness (per subject)     & $\mathcal{A}_{sujet}$  \\
                               &                        \\
ideal/pleasant                & i                      \\
non-ideal/unpleasant         & ni                     \\
scene ideal/pleasant          & i-scene                \\
scene non-ideal/unpleasant   & ni-scene               \\
\hline
\end{tabular}
\vspace{0.5mm}
\caption{Acronyms of features used in the analysis of the sensory experiments.}
\label{tab:acronyme}
\end{table}

@@

\begin{itemize}
\item \emph{perceptual features}: the perceived pleasantness of the composed scene, assessed on a 7-point Likert scale. $\mathcal{A}_{scene}$ is the average pleasantness for one scene, as the average of the marks of all the participants. $\mathcal{A}_{subject}$ is the average pleasantness for each participant, computed as the average of all the score given by each participant. Given the low number of participants in experiment 1.b, we choose not to normalize the pleasantness scores.

\item \emph{semantic features}: a boolean vector $S = (x_1, x_2, \ldots, x_n)$ that indicates the class of sounds involved in the scene. Each boolean $x$ of this vector corresponds to a specific class of sounds: $x = 1$ if the class is present in the scene, and $x = 0$ otherwise. The vector dimension ($n$) depends on the level of abstraction that is considered for analysis. For instance, for the abstraction level 1 that includes 44 classes of sounds, $n = 44$.

\item \emph{structural features}: features related to the \emph{sound level}. To figure out the sound level, we take the $L_{Aeq}$ feature as a conceptual reference. In our case, it is a scalar computed from the  signal (in Volt and not in Pascal), and converted in decibels, with a reference of 1 Volt (full scale). The level is obtained by computing the quadratic mean of the signal every second and averaging the results over the total duration of the scene. An A-filtering module processes the data before the quadratic means are computed. We note $L$, $L(E)$ and $L(T)$ the computed levels by respectively considering the whole set of samples, only the set of event samples and only the set of texture samples.
\end{itemize}

@@

\item \emph{analysis of the semantic features}: the goal is to evaluate if specific classes are more frequently used to simulate a given type of environment (i or ni). A V-test is considered \citep{lebart2000statistique} \ml{y pas une ref en Anglais ?}. With $c$ being the total number of classes used for both types of environment, $c_k$ the number of classes used for a given type of scene $k$ ($k=\lbrace i,ni\rbrace$), $c_j$ the number of time a class $j$ has been used, and $c_{jk}$ the number of time a class $j$ has been used for a given type of scene $k$, the V-test tests the null hypothesis that the ratio $\frac{c_{jk}}{c}$ is not significatively different from the ratio $\frac{c_{jk}}{c_k}$. For each class $j$, and each environment type $k$, an approximation of the statistic value $V_{jk}$ is computed as follows:

@@

If the hypothesis is rejected, the class $j$ is said to be typical with respect to the type of scene $k$. The typical classes are called \textbf{sound markers}. Testing is done for each class, at every level of abstraction, and separately for texture and event classes;

@@

First, in order to ensure the coherence of the data, we check that none of the ni-scenes gets a $\mathcal{A}_{scene}$ higher than one of a i-scene. Four ni-scenes don’t fullfill that constraint: they are thus removed, together with their corresponding i-scenes. As a consequence, 36 i-scenes and 36 ni-scenes remain for analysis. Second, we verify that participants really perceived a difference in terms of pleasantness between i- and ni-scenes. For that, we investigate the mean pleasantness score for each participant $\mathcal{A}_{subject}$, computed separately for each type of environment. It appears that the i-scenes were perceived significatively more pleasant ($p < 0.01$) than the ni-scenes.

@@

First, our analyses focuses on the sound level. Figures ~\ref{fig:soundlevela},~\ref{fig:soundlevelb} and~\ref{fig:soundlevelc} respectively depict the levels $L$, $L(E)$ and $L(T)$. There is a significative difference in terms of sound level between i- and ni-scenes ($L$: $p<0.01$). This difference is siginificant for events ($L(E)$: $p<0.01$) and for textures ($L(T)$: $p<0.01$).
%, with a standard deviation of -7 $dB$
%, STD: -7 $dB$
%, STD: -6 $dB$

@@

It appears that the biggest influence on the global sound level comes from the events, the gap between $L$ and $L(E)$ being only 1 $dB$ for i- and ni-scenes. This observation is in agreement with the results obtained by Kuwano~\al~\cite{kuwano_memory_2003}. During their experiment, the authors ask their participants to assess a set of soundscapes at a global level and then to do the same judgment at the time when they detect a sound source. The study shows that there is no significative difference between global and averaged instantaneous judgments. In our case, the result can be interpreted as if the participants had unconsciously integrated this perceptual reality when composing the scenes, by assigning the global sound level to well identified short sounds, \ie the events.

@@

\begin{figure*}
        \subfloat[]
        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_1}\label{fig:soundlevela}}
        \subfloat[]
        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_3}\label{fig:soundlevelb}}
        \subfloat[]
        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_5}\label{fig:soundlevelc}}\par
        \subfloat[]
        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_2}\label{fig:soundleveld}}
        \subfloat[]
        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_4}\label{fig:soundlevele}}
        \subfloat[]
        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_6}\label{fig:soundlevelf}}
       \caption{Distributions of the structural features $L$ (a, d), $L(E)$ (b, e) et $L(T)$ (c, f), with respect to scene type (a, b, c) and perceived pleasantness $\mathcal{A}_{scene}$ of experiments 1.b (d, e, f).}
\end{figure*}

@@

In this section, relationships that could exist between structural features and perceived pleasantness are investigated. Contrary to the previous section, we do not consider the pleasantness property of a scene on a binary mode (i vs. ni): this is the mean pleasantness $\mathcal{A}_{scene}$ that is considered here as a perceptual feature. The aim is to investigate the level of correlation between structural features and $\mathcal{A}_{scene}$. The linear correlation coefficients computed between $\mathcal{A}_{scene}$ \vs~$L$, $L(E)$, $L(T)$ are shown on Table~\ref{tab:corrStructA}. Relationships between $\mathcal{A}_{scene}$ and the structural features are depicted in Figure~\ref{fig:soundleveld},~\ref{fig:soundlevele} and ~\ref{fig:soundlevelf}.  \jfp{en table 2, il devrait y avoir les coefficients de corrélation pour toutes les données, puis pour les deux cas ni et i.
}

@@

\begin{table}[t]
\centering
\begin{tabular}{l c c c}
               & i-scenes                   & ni-scenes    \\
\hline
$L$            & -0.32 ($p=0.06$)           & \textbf{-0.78} ($p<0.01$)\\
$L(E)$         & -0.20 ($p=0.24$)           & \textbf{-0.75} ($p<0.01$)\\
$L(T)$         & -0.33 ($p=0.05$)           &  -0.00 ($p=0.99$) \\
\hline
\end{tabular}
\vspace{0.5mm}
\caption{Linear correlation coefficients computed between mean perceived pleasantness $\mathcal{A}_{scene}$ of experiment 1.b and structural features.\jfp{rajouter une colonne corélation globale}}
\label{tab:corrStructA}
\end{table}

@@

@@

\begin{figure}
        \subfloat[]
        {\includegraphics[width=1\linewidth]{gfx/ch_5/xp1_class_1}\label{fig:soundsourcea}} \par
        \subfloat[]
        {\includegraphics[width=.5\linewidth]{gfx/ch_5/xp1_class_2}\label{fig:soundsourceb}}
        \subfloat[]
        {\includegraphics[width=.5\linewidth]{gfx/ch_5/xp1_class_3}\label{fig:soundsourcec}}
       \caption{Proportion of synthesized scenes (i or ni) involving a specific class of sounds: (a) event classes at an abstraction level $0+$, (b) texture classes at an abstraction level $0$, (c) event sub-classes at an abstraction level $1$ belonging to \emph{traffic} and \emph{public transportation} classes of the abstraction level $0$.}\label{fig:soundsource}
\end{figure}

@@

We highlight that, on a qualitative point of view, the composition of the scenes in terms of sound sources differs with respect to the type of environment (i or ni). Let us now investigate whether some of the sound classes are specific to a given environment. For that, we use the V-test as detailed in  Section~\ref{sec:xp1_dataAna} separately for each abstraction level. Results are presented in Table~\ref{tab:markers}. \jfp{le V-test est un test statistique : on peut mettre la p-value dans le tableau}

@@

\begin{table*}[t]
 \setlength{\tabcolsep}{0.2pt}
 \centering
  {\renewcommand{\arraystretch}{0.9}
\begin{tabular}{c c c c}
Abstraction        & \multicolumn{2}{c}{Event sound markers} \\
level & i-scenes & ni-scenes \\
\hline
0  &                          & construction work (3.78)  \\
\hline
  & church bell  (4.5)             & car horn  (3.9) \\
1 & bicycle bell  (4.3)      & siren (3.9)\\
  & animal (4.2)              &       \\
   \hline
  & bird        (4.8)       & car horn  (4.0)\\
2 & church bell  (4.4)             & siren (4.0)\\
  & bicycle bell     (4.2)             &       \\
   \hline
  & bird song (4.8)        & klaxon  (4.1)\\
3 & church bell   (4.3)            & siren (4.0)\\
  & bicycle bell      (4.2)   &       \\
  & foot steps  (3.6)      &  \\
  &                           & \\
  & \multicolumn{2}{c}{Texture sound markers}      \\
  & i-scenes & ni-scenes \\
\hline
0 &     courtyard / park (4.1) &  construction (3.9)  \\
\hline
1 &     park (3.65)          &  crossroad (3.6)  \\
  &                          &  construction vehicule (3.3)  \\
\hline
2 &     park (3.64)          &  crossroad (3.56)  \\
\hline
\end{tabular}
}
\vspace{0.5mm}
\caption{Event and texture classes identified as sound markers. In each cell, markers are ranked in the decreasing order of V-test value. $p\leq0.01$ for all sound markers.}
\label{tab:markers}
\end{table*}

@@

@@

\begin{enumerate}
\item a significative difference between levels of i- and ni-scenes exists ($L_m$, $L(E)_m$ et $L(T)_m$: $p<0.01$);
\item the sound level of scenes is mainly related to the sound events, compared to the textures;
\item the sound level of events has an influence on the perception of pleasantness for ni-scenes, but not for i-scenes;
\item the sound level of textures does not play any role in the perception of the pleasantness.
\end{enumerate}

@@

Besides, whereas $L(T)$ did not show any significative correlation for ni-scenes, a strong negative correlation is observed for $L(T)_b$ ($r = -0.73$, $p < 0.01$). This indicates that the level of non-markers texture classes does not influence the perceived pleasantness in the same way for i- and ni-scenes. Sound level seems to have a negative effect for the ni-scenes, and no significant effect for the i-scenes.

@@

For the ni-scenes, the fact that we observe significative correlations for $L_m$ and $L(E)_m$ and no correlation for $L_m-L_b$ and $L(E)_m$ - $L(E)_b$ shows that this is indeed the absolute level that matters.

\begin{figure*}[t]
        \myfloatalign
        \subfloat[]
        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_7}\label{fig:soundlevelMarkera}}
        \subfloat[]
        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_9}\label{fig:soundlevelNoisea}}
        \subfloat[]
        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_19}\label{fig:soundlevelMarkerDiffa}}\par
        \subfloat[]
        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_8}\label{fig:soundlevelMarkerd}}
        \subfloat[]
        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_10}\label{fig:soundlevelNoised}}
        \subfloat[]
        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_20}\label{fig:soundlevelMarkerDiffd}}
        \caption{Distribution of the relative sound levels structural features related to the presence of markers $L_m$ (a, d), $L_b$ (b, e) and $L_m-L_b$ (c, f), versus scene type (a, b, c) and perceived pleasantness $\mathcal{A}_{scene}$ of experiment 1.b (d, e, f).}\label{fig:soundlevelMarker}
\end{figure*}

@@

%\jfp{on se demande pourquoi on va de -5 a +5 alors qu'avant on allait de -3 à +3}

@@

12 subjects (4 women) perform the test. None of them have performed experiments 1.a and 1.b. The subjects are 22 to 61 years old (average: 29.5, standard deviation: 14). All the subjects live in Nantes, France, for at least two years. All the subjects performed the test successfully.

@@

@@

Let us consider $\mathcal{A}_{sujet}$ for the km-scenes. Close inspection of the answers show that one subject strongly differs from the ones of others. This subject evaluated positively close to half of the ni/km-scenes, see Annex~\ref{app:xp2}, subject 7. This subject also gave a score above 0 for 58\% of the ni/km-scenes, contrary to an average of 11\% for the remaining of the subjects.

@@

In this section, we study the scores given by the subject while listening to the several types of scenes, namely the i/km-scenes, ni/km-scenes, i/rm-scenes and ni/rm-scenes. The repeated measure ANOVA applied to $\mathcal{A}_{sujet}$ show a significative effect of the type environment (i/ni: $F[1,10]=175$, $p<0.01$), of the presence/absence of markers (km/rm: $F[1,10]=7$, $p<0.05$), and of the interaction between those two factors ($F[1,10]=67$, $p<0.01$).

>>>>

\newcommand{\gl}[1]{\textcolor{magenta}{GL : #1}}
\newcommand{\mr}[1]{\textcolor{green}{MR : #1}}
\newcommand{\ml}[1]{\textcolor{red}{ML : #1}}
\newcommand{\jfp}[1]{\textcolor{blue}{JFP : #1}}
\newcommand{\ie}{\emph{i.\,e.}}
\newcommand{\Ie}{\emph{I.\,e.}}
\newcommand{\cf}{cf.}
\newcommand{\Cf}{Cf.}
\newcommand{\vs}{\emph{vs.}}
\newcommand{\Vs}{\emph{Vs.}}
\newcommand{\al}{\emph{et~al.}}
\newcommand{\eg}{\emph{e.\,g.}}
\newcommand{\Eg}{\emph{E.\,g.}}
\newcommand{\etc}{\emph{etc.}}
\newcommand{\myfloatalign}{\centering}

@@

  \begin{abstract}
    This paper introduces a new experimental protocol to study mental representations of urban soundscapes through a simulation process. Subjects are asked to create a full soundscape by means of a dedicated software tool, coupled with a structured sound data set. This paradigm is used to characterize urban sound environment representations by analyzing the sound classes that were used to simulate the auditory scenes. A rating experiment of the soundscape pleasantness using a 7-point bipolar semantic scale is conducted to further refine the analysis of the simulated urban acoustic scenes. Results show that 1) a semantic characterization in terms of presence~/ absence of sound sources is an effective way to characterize urban soundscapes pleasantness, and 2) physical descriptors computed for specific sound sources better characterize the appraisal than global descriptors.
  \end{abstract}

  \begin{keyword}
  %% keywords here, in the form: keyword \sep keyword
 cognitive psychology \sep soundscape perception \sep soundscape simulator
  %% PACS codes here, in the form: \PACS code \sep code

@@

\begin{figure}[t]
        \myfloatalign
        \includegraphics[trim={0 0 0 2.5cm},clip,width=.8\linewidth]{gfx/ch_4/3.png}
       \caption{Hierarchical organisation of the isolated sounds used in the simulation.}\label{fig:orgDb}
\end{figure}

@@

\begin{figure}[!tp]
		\myfloatalign
        \subfloat[]
        {\includegraphics[width=.8\linewidth]{gfx/ch_4/SSF}\label{fig:ssf}} \par
        \subfloat[]
        {\includegraphics[width=\linewidth]{gfx/ch_4/simscene}\label{fig:simscene}}
       \caption{Graphical interfaces for the selection of sound classes (a) and their sequencing (b) of the \emph{simScene} tool.}
\end{figure}

@@

\begin{figure}[t]
        \includegraphics[width=\linewidth]{gfx/ch_5/5.png}
        \caption{Experimental protocol of the simulation experiment (1.a) and the pleasantness evaluation experiment (1.b)}\label{fig:xp1_2}
\end{figure}

@@

44 students (14 females; averaging $21.6$ years of age, s.d. of $2.0$ years) from \emph{Ecole Centrale de Nantes} (a french engineering school) took part to the experiment. All the subjects live in Nantes, France, for at least two years

@@

The assessment is done with a 7-point bipolar semantic scale going from -3 (non ideal / unpleasant) to +3 (ideal / pleasant). Before evaluating a scene, participants must listen to the first 20 seconds of the stimuli. After the evaluation, they are free to switch to the next scene.

@@

10 students (2 females; averaging $23.1$ years of age, s.d. of $1.8$ years) from \emph{Ecole Centrale de Nantes} (a french engineering school) took part to the experiment. All the subjects live in Nantes, France, for at least two years. None of them took part to the previous simulation experiment (experiment 1.a).

@@

\begin{table}[t]
\centering
\begin{tabular}{c c}
Terms                              & Acronyms                   \\
\hline
Sound level                        & $L$                        \\
Sound level (events)               & $L(E)$                     \\
Sound level (textures)             & $L(T)$                     \\
Average pleasantness (per scene)   & $\mathcal{A}_{scene}$      \\
Average pleasantness (per subject) & $\mathcal{A}_{subject}$      \\
                                   &                            \\
ideal/pleasant                     & i                          \\
non-ideal/unpleasant               & ni                         \\
scene ideal/pleasant               & i-scene                    \\
scene non-ideal/unpleasant         & ni-scene                   \\
\hline
\end{tabular}
\vspace{0.5mm}
\caption{Acronyms of features used in the analysis of the sensory experiments.}
\label{tab:acronyme}
\end{table}

@@

\begin{itemize}
\item \emph{perceptual features}: the perceived pleasantness of the composed scene, assessed on a 7-point bipolar semantic scale. $\mathcal{A}_{scene}$ is the average pleasantness for one scene, as the average of the marks of all the participants. $\mathcal{A}_{subject}$ is the average pleasantness for each participant, computed as the average of all the score given by each participant. Given the low number of participants in experiment 1.b, we choose not to normalize the pleasantness scores.

\item \emph{semantic features}: the sound classes that are present / absent from the scene. We use a boolean vector $S = (x_1, x_2, \ldots, x_n)$ that indicates the class of sounds involved in the scene. Each boolean $x$ of this vector corresponds to a specific class of sounds: $x = 1$ if the class is present in the scene, and $x = 0$ otherwise. The vector dimension ($n$) depends on the level of abstraction that is considered for analysis. For instance, for the abstraction level 1 that includes 44 classes of sounds, $n = 44$.

\item \emph{structural features}: the \emph{sound level}. To figure out the sound level, we draw inspiration from the $L_{Aeq}$ feature. In our case, it is a scalar computed from the  signal (in Volt and not in Pascal), and converted in decibels, with a reference of 1 Volt (full scale). The level is obtained by computing the quadratic mean of the signal every second and averaging the results over the total duration of the scene. An A-filtering module processes the data before the quadratic means are computed. We note $L$, $L(E)$ and $L(T)$ the computed levels by respectively considering the whole set of samples, only the set of event samples and only the set of texture samples.
\end{itemize}

@@

\item \emph{analysis of the semantic features}: the goal is to evaluate if specific classes are more frequently used to simulate a given type of environment (i or ni). A V-test is considered \citep{lebart2000statistique} \ml{ref en Anglais ?}. With $c$ being the total number of classes used for both types of environment, $c_k$ the number of classes used for a given type of scene $k$ ($k=\lbrace i,ni\rbrace$), $c_j$ the number of time a class $j$ has been used, and $c_{jk}$ the number of time a class $j$ has been used for a given type of scene $k$, the V-test tests the null hypothesis that the ratio $\frac{c_{jk}}{c}$ is not significantly different from the ratio $\frac{c_{jk}}{c_k}$. For each class $j$, and each environment type $k$, an approximation of the statistic value $V_{jk}$ is computed as follows:

@@

If the null hypothesis is rejected, the class $j$ is said to be typical with respect to the type of scene $k$. The typical classes are called \textbf{sound markers}. Testing is done for each class, at every level of abstraction, and separately for texture and event classes;

@@

First, in order to ensure the coherence of the data, we check that none of the ni-scenes gets a $\mathcal{A}_{scene}$ higher than one of a i-scene. 4 ni-scenes don't fulfill that constraint: they are thus removed, together with their corresponding i-scenes. As a consequence, 36 i-scenes and 36 ni-scenes remain for analysis. Second, we verify that participants really perceived a difference in terms of pleasantness between i- and ni-scenes. For that, we investigate the mean pleasantness score for each participant $\mathcal{A}_{subject}$, computed separately for each type of environment. It appears that the i-scenes were perceived significantly more pleasant ($p < 0.01$) than the ni-scenes.

@@

\begin{figure}[t]
		\myfloatalign
        \subfloat[]
        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_1}\label{fig:soundlevela}}
        \subfloat[]
        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_3}\label{fig:soundlevelb}}
        \subfloat[]
        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_5}\label{fig:soundlevelc}}\par
        \subfloat[]
        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_2}\label{fig:soundleveld}}
        \subfloat[]
        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_4}\label{fig:soundlevele}}
        \subfloat[]
        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_6}\label{fig:soundlevelf}}
       \caption{Distributions of the structural features $L$ (a, d), $L(E)$ (b, e) et $L(T)$ (c, f), with respect to scene type (a, b, c) and perceived pleasantness $\mathcal{A}_{scene}$ of experiments 1.b (d, e, f).}
\end{figure}

First, our analyses focuses on the sound level. Figures ~\ref{fig:soundlevela},~\ref{fig:soundlevelb} and~\ref{fig:soundlevelc} respectively depict the levels $L$, $L(E)$ and $L(T)$. There is a significant difference in terms of sound level between i- and ni-scenes ($L$: $p<0.01$). This difference is siginificant for events ($L(E)$: $p<0.01$) and for textures ($L(T)$: $p<0.01$).
%, with a standard deviation of -7 $dB$
%, STD: -7 $dB$
%, STD: -6 $dB$

@@

It appears that the biggest influence on the global sound level comes from the events, the gap between $L$ and $L(E)$ being only 1 $dB$ for i- and ni-scenes. This observation is in agreement with the results obtained by Kuwano~\al~\cite{kuwano_memory_2003}. During their experiment, the authors ask their participants to assess a set of soundscapes at a global level and then to do the same judgment at the time when they detect a sound source. The study shows that there is no significant difference between global and averaged instantaneous judgments. In our case, the result can be interpreted as if the participants had unconsciously integrated this perceptual reality when composing the scenes, by assigning the global sound level to well identified short sounds, \ie the events.

@@

@@

\begin{table}[t]
\centering
\begin{tabular}{l c c c}
               & all scenes                & i-scenes          & ni-scenes    \\
\hline
$L$            & \textbf{-0.77} ($p<0.01$) & -0.32 ($p=0.06$)  & \textbf{-0.78} ($p<0.01$)\\
$L(E)$         & \textbf{-0.75} ($p<0.01$) & -0.20 ($p=0.24$)  & \textbf{-0.75} ($p<0.01$)\\
$L(T)$         & \textbf{-0.53} ($p<0.01$) & -0.33 ($p=0.05$)  &  -0.00 ($p=0.99$) \\
\hline
\end{tabular}
\vspace{0.5mm}
\caption{Linear correlation coefficients computed between mean perceived pleasantness $\mathcal{A}_{scene}$ of experiment 1.b and structural features.}
\label{tab:corrStructA}
\end{table}

In this section, relationships that could exist between structural features and perceived pleasantness are investigated. Contrary to the previous section, we do not consider the pleasantness property of a scene on a binary mode (i vs. ni): this is the mean pleasantness $\mathcal{A}_{scene}$ that is considered here as a perceptual feature. The aim is to investigate the level of correlation between structural features and $\mathcal{A}_{scene}$. The linear correlation coefficients computed between $\mathcal{A}_{scene}$ \vs~$L$, $L(E)$, $L(T)$ are shown on Table~\ref{tab:corrStructA}. Relationships between $\mathcal{A}_{scene}$ and the structural features are depicted in Figure~\ref{fig:soundleveld},~\ref{fig:soundlevele} and~\ref{fig:soundlevelf}.

@@

@@

\begin{figure}[!tp]
		\myfloatalign
        \subfloat[]
        {\includegraphics[width=1\linewidth]{gfx/ch_5/xp1_class_1}\label{fig:soundsourcea}}\par
        \subfloat[]
        {\includegraphics[width=.5\linewidth]{gfx/ch_5/xp1_class_2}\label{fig:soundsourceb}}
        \subfloat[]
        {\includegraphics[width=.5\linewidth]{gfx/ch_5/xp1_class_3}\label{fig:soundsourcec}}
       \caption{Proportion of simulated scenes (i or ni) involving a specific class of sounds: (a) event classes at an abstraction level $0+$, (b) texture classes at an abstraction level $0$, (c) event sub-classes at an abstraction level $1$ belonging to \emph{traffic} and \emph{public transportation} classes of the abstraction level $0$.}\label{fig:soundsource}
\end{figure}

@@

@@

\begin{table}[t]
 \setlength{\tabcolsep}{0.2pt}
 \centering
  {\renewcommand{\arraystretch}{0.9}
\begin{tabular}{c c c c}
Abstraction        & \multicolumn{2}{c}{Event sound markers} \\
level & i-scenes & ni-scenes \\
\hline
0  &                          & construction work (3.78)  \\
\hline
  & church bell  (4.5)             & car horn  (3.9) \\
1 & bicycle bell  (4.3)      & siren (3.9)\\
  & animal (4.2)              &       \\
   \hline
  & bird        (4.8)       & car horn  (4.0)\\
2 & church bell  (4.4)             & siren (4.0)\\
  & bicycle bell     (4.2)             &       \\
   \hline
  & bird song (4.8)        & klaxon  (4.1)\\
3 & church bell   (4.3)            & siren (4.0)\\
  & bicycle bell      (4.2)   &       \\
  & foot steps  (3.6)      &  \\
  &                           & \\
  & \multicolumn{2}{c}{Texture sound markers}      \\
  & i-scenes & ni-scenes \\
\hline
0 &     courtyard / park (4.1) &  construction (3.9)  \\
\hline
1 &     park (3.65)          &  crossroad (3.6)  \\
  &                          &  construction vehicule (3.3)  \\
\hline
2 &     park (3.64)          &  crossroad (3.56)  \\
\hline
\end{tabular}
}
\vspace{0.5mm}
\caption{Event and texture classes identified as sound markers. In each cell, markers are ranked in the decreasing order of V-test value. $p\leq0.01$ for all sound markers.}
\label{tab:markers}
\end{table}

We highlight that, on a qualitative point of view, the composition of the scenes in terms of sound sources differs with respect to the type of environment (i or ni). Let us now investigate whether some of the sound classes are specific to a given environment. For that, we use the V-test as detailed in  Section~\ref{sec:xp1_dataAna} separately for each abstraction level. Results are presented in Table~\ref{tab:markers}.

@@

@@

\begin{figure}[t]
        \myfloatalign
        \subfloat[]
        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_7}\label{fig:soundlevelMarkera}}
        \subfloat[]
        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_9}\label{fig:soundlevelNoisea}}
        \subfloat[]
        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_19}\label{fig:soundlevelMarkerDiffa}}\par
        \subfloat[]
        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_8}\label{fig:soundlevelMarkerd}}
        \subfloat[]
        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_10}\label{fig:soundlevelNoised}}
        \subfloat[]
        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp_soundlevel_20}\label{fig:soundlevelMarkerDiffd}}
        \caption{Distribution of the relative sound levels structural features related to the presence of markers $L_m$ (a, d), $L_b$ (b, e) and $L_m-L_b$ (c, f), versus scene type (a, b, c) and perceived pleasantness $\mathcal{A}_{scene}$ of experiment 1.b (d, e, f).}\label{fig:soundlevelMarker}
\end{figure}

@@

\begin{enumerate}
\item a significant difference between levels of i- and ni-scenes exists ($L_m$, $L(E)_m$ et $L(T)_m$: $p<0.01$);
\item the sound level of scenes is mainly related to the sound events, compared to the textures;
\item the sound level of events has an influence on the perception of pleasantness for ni-scenes, but not for i-scenes;
\item the sound level of textures does not play any role in the perception of the pleasantness.
\end{enumerate}

@@

Besides, whereas $L(T)$ did not show any significant correlation for ni-scenes, a strong negative correlation is observed for $L(T)_b$ ($r = -0.73$, $p < 0.01$). This indicates that the level of non-markers texture classes does not influence the perceived pleasantness in the same way for i- and ni-scenes. Sound level seems to have a negative effect for the ni-scenes, and no significant effect for the i-scenes.

@@

For the ni-scenes, the fact that we observe significant correlations for $L_m$ and $L(E)_m$ and no correlation for $L_m-L_b$ and $L(E)_m$ - $L(E)_b$ shows that this is indeed the absolute level that matters.

@@

@@

12 subjects perform the test (4 females; averaging $29.5$ years of age, s.d. of $14$ years). None of them have performed experiments 1.a and 1.b. All the subjects live in Nantes, France, for at least two years. All the subjects performed the test successfully.

@@

\begin{figure}[t]
        \myfloatalign
        \subfloat[]
        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp4_note_13}\label{fig:xp2_Asubject}}
        \subfloat[]
        {\includegraphics[width=.33\linewidth]{gfx/ch_5/xp4_note_15}\label{fig:xp2_Ascene}}
        \caption{Distribution of the mean perceived pleasantness per subject $\mathcal{A}_{subject}$ (a) and the mean perceived pleasantness  per scenes$\mathcal{A}_{scene}$ (b) versus scene type. Black stars in subfigure (a) stand for the detected outlier,~\ie~subject 7.}\label{fig:xp2A}
\end{figure}

@@

Let us consider $\mathcal{A}_{subject}$ for the km-scenes, see Figure~\ref{fig:xp2_Asubject}. Close inspection of the answers show that one subject strongly differs from the ones of others. This subject evaluated positively close to half of the ni/km-scenes, see Annex~\ref{app:xp2}, subject 7. This subject also gave a score above 0 for 58\% of the ni/km-scenes, contrary to an average of 11\% for the remaining of the subjects.

@@

In this section, we study the scores given by the subject while listening to the several types of scenes, namely the i/km-scenes, ni/km-scenes, i/rm-scenes and ni/rm-scenes, see Figure~\ref{fig:xp2_Ascene}. The repeated measure ANOVA applied to $\mathcal{A}_{subject}$ show a significant effect of the type environment (i/ni: $F[1,10]=175$, $p<0.01$), of the presence/absence of markers (km/rm: $F[1,10]=7$, $p<0.05$), and of the interaction between those two factors ($F[1,10]=67$, $p<0.01$).
